Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-10-31 workshop-paper/paper.tex
Macros
* Usage of macros for IW, BFS and UCT enforced

\newcommand{\propername}[1]{\textsc{#1}}
\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
%\newcommand{\IW}{{\textit{IW}}}
%\newcommand{\SIW}{{\textit{SIW}}}
%\newcommand{\ID}{{\textit{ID}}}
%\newcommand{\BRFS}{{\textit{BrFS}}}
%\newcommand{\BFS}{{\textit{2BFS}}}
%\newcommand{\UCT}{{\textit{UCT}}}
\newcommand{\IW}{{\textsc{IW}}}
\newcommand{\SIW}{{\textsct{SIW}}}
\newcommand{\ID}{{\textsc{ID}}}
\newcommand{\BRFS}{{\textsc{BrFS}}}
\newcommand{\BFS}{{\textsc{2BFS}}}
\newcommand{\UCT}{{\textsc{UCT}}}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly,  the action effects and goals are  not known a priori. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or IW, consists of  a sequence of calls  IW(1), IW(2), .., IW(k) where  IW(i) is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of IW with the  $k$ parameter fixed to $1$, 
i.e., IW(1), is much better than random or breadth-first search, and approaches 
the performance of UCT. A simple best-first variation  of IW,  that combine exploration and exploitation,
is evaluated as well. 
\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and 
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}. 
Results have been reported so far for basic planning algorithms like breadth-first 
search and UCT \cite{bellemare:jair2013}, and for  reinforcement learning
and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the 
performance of  human players across a broad range of games.

@@

The presence of unknown transition and  rewards in  the Atari games does not preclude the use of blind-search
methods like breadth-first search, Dijkstra's algorithm \cite{dijkstra},  or learning methods  such as 
LRTA* \cite{korf:lrta}, UCT \cite{uct}, and Q-learning  \cite{sutton:rl,bertsekas:neuro}. Indeed, 
the net-benefit planning problem with unknown state transitions and rewards over a given planning
horizon, can be mapped into a standard \emph{shortest-path problem} which can be solved optimally by Dijkstra's algorithm.
For this, we just need to  map the unknown rewards $r(a,s)$ into positive (unknown) action costs $c(a,s) = C - r(a,s)$
where  $C$ is a large constant that exceeds the maximum possible reward. The fact that the state 
transition and cost functions  $f(a,s)$ and $c(a,s)$ are not known a priori doesn't affect the applicability of 
Dijkstra's algorithm, which requires the value of these functions precisely when  the action $a$
is applied in the state $s$.

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line. 
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planning algorithms: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is 
called IW for Iterated Width search \cite{nir:ecai2012}. IW consists of a sequence of calls  IW(1), IW(2), .., IW(k), where 
IW(i) is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set) 
of at most $i$ atoms. Namely,   IW(1) is a breadth-first search that keeps  a state only if  the state is   the first  one 
in the search to make some atom  true; IW(2) keeps a state  only if the state is  the first one to  make  a pair of atoms true, and 
so on.  Like plain breadth-first  and iterative deepening searches,  IW  is 
complete, while searching the state space in a way that makes use of the 
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables 
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$ 
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$, 
the procedure $IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$. 
Elsewhere, it has been shown that  IW exhibits   state-of-the-art performance on most existing planning benchmark domains 
when goals are restricted to be single atoms  where IW provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a block world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time  
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that IW(2) will be 
not only complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for any instance, as long as the goal is atomic. 
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built, 
simple extensions of the basic IW procedure have been developed as well, such as Serialized IW, where IW is 
used to achieve the goals one at at time \cite{nir:ecai2012,nir:ecai2014}. 
The extension of IW for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

The Iterative Width  (IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the state space, a problem is defined by an initial state $s_0$, 
a set of actions applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not  need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

IW consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached. 
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state 
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting, 
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by IW is the path  that has a  maximum accumulated reward. 
The accummulated reward $R(s)$  of a state $s$ reached in an iteration of IW is determined by the  unique parent state $s'$ 
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is then the first action along such a path.

@@

\begin{table}[t!]
\begin{center}
%\resizebox{3in}{!}{
{\scriptsize
% \begin{tabular}{r@{}lrrrr@{}}
\begin{tabular}{llcccc}
%% table header
\hline\\[-1.3ex]
%\toprule
\# & \multicolumn{1}{c}{Domain} & \multicolumn{1}{c}{I} & \multicolumn{1}{c}{IW(1)} & \multicolumn{1}{c}{IW(2)} & \multicolumn{1}{c}{Neither} \\[.2ex]

@@

\caption{\small 
\emph{Top:} Number of classical planning problems  and percentages solved by IW(1), IW(2), or neither. Instances obtained
by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),  
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes and 2 GB resp. From \cite{nir:ecai2102}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

IW is a systematic and complete blind-search algorithm like breadth-first search (BrFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very 
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition. 
Yet, this changes the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of IW, i.e.  IW(1), by the second iteration 
IW(2), and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by IW(1) while $51.3\%$ are  solved by IW(2). Since IW($k$)
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encodings are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$ 
\cite{bonet:hsp-aij} on the other (this algorithm is similar 
to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34627 vs. 34849, far ahead
of the other blind algorithms BrFS and ID  that solve 9010 and 8762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $IW(k)$ solves a problem  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{IW runs in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  
relevant  in  optimization problems as well. Indeed, if a  good path is made of 
states $s_i$ each of which have a low width, IW can be made 
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in IW to enforce this property
which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by IW(1) is $n \times D \times b$ in the worst case, 
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of 
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by IW(2) in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of 
such states. For classical planning problems, the growth in the number of nodes from 
IW(1) to IW(2) is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables, 
and apply these algorithms to that representation. The results would be different.
In such a case, IW(1) and IW(2)  would generate $n' \times D' \times b$ and 
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't tried  this representation yet
under the assumption that the correlations among the bits in each one of 
the 128 words of the state vectors are meaningful. 
In summary, from the basic IW algorithm we are testing only the first 
call IW(1).

IW is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd,malte:alternating}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with no accumulated reward  that are 
placed in the first queue only. We refer to this best-first algorithm 
as 2BFS.

For  the experiments below, we added two simple variations to  IW(1) and 2BFS.
First, in the breadth-first search underlying IW(1), we generate the children in 
random order. This makes the executions that results from the
IW(1) lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms 
for discounting future rewards like in UCT. 
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$, 
$R(s')$ is set to $R(s) + \gamma^d(s´) r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier, 
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

We tested  $\IW(1)$ and \BFS\  over $9$ games: those considered in the \emph{training} set
of \cite{bellemare:jair2013} and those discussed in \cite{deep-mind-atari}. 
In Table~\ref{performance} the performance of  $\IW(1)$ and \BFS\ 
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead, 
is compared with that of \UCT, breadth-first search (\BRFS), and random action selection. 
Our experimental setup  follows the one in  \cite{bellemare:jair2013} with two exceptions. First, rather than
considering $10$ episodes of each game, we have considered just $5$ due to time constraints, as the experiments 
for each  game can take up to $40$ hours of CPU time. Second, while \UCT\ is still limited to perform up to $500$ base policy rollouts 
of depth $300$, a maximum of budget of  $100,000$ simulated frames is applied to all the algorithms, UCT included. 
%% Was not clear; hope this is what you meant, if not fix
%%% ($150,000$ frames are to be simulated while performing the lookahead), both $\IW(1)$ and \BFS are limited to
% The other \UCT\ parameters like the exploration constant $C$ and the horizon $m$ are the same ($C=0.1$, $m=300$). 
%% matching the setup used during actual gameplay. %%  (an action is selected every $5$ frames, using a no-op for the remaining $4$). %% Needed?
The average scores reported for \UCT\  \cite{bellemare:jair2013}, without this extra limitation, are included
in  parenthesis for reference. $\IW(1)$ and \BFS\ have been  limited to a planning horizon $m=1500$ where one an action is evaluated every 
$5$ frames. The discount factor used by all the algorithms is $\gamma=0.995$, and all of them  reuse the tree built 
in the previous lookahead by setting the root of the tree to the best child of the root, while 
deleting its siblings and their descendants as in \cite{bellemare:jair2013}.
%%% really need what's below? it's not very clear .. at least to me HG
%% We note that  this may have a different effect on algorithms where all branches have the same number of
%% generated nodes, as is the case of \BRFS, than in $\IW(1)$, \BFS\ and \UCT, where the depth of
%% the branches is uneven.

>>>>

\newcommand{\propername}[1]{\textsc{#1}}
\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
%\newcommand{\IW}{{\textit{IW}}}
%\newcommand{\SIW}{{\textit{SIW}}}
%\newcommand{\ID}{{\textit{ID}}}
%\newcommand{\BRFS}{{\textit{BrFS}}}
%\newcommand{\BFS}{{\textit{\BFS}}}
%\newcommand{\UCT}{{\textit{UCT}}}
\newcommand{\IW}{{\textsc{IW}}}
\newcommand{\SIW}{{\textsct{SIW}}}
\newcommand{\ID}{{\textsc{ID}}}
\newcommand{\BRFS}{{\textsc{BrFS}}}
\newcommand{\BFS}{{\textsc{2BFS}}}
\newcommand{\UCT}{{\textsc{UCT}}}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly,  the action effects and goals are  not known a priori. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, 
.., $\IW(k)$ where  $\IW(i)$ is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$, 
i.e., $\IW(1)$, is much better than random or breadth-first search, and approaches 
the performance of \UCT. A simple best-first variation  of \IW,  that combine exploration and exploitation,
is evaluated as well. 
\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and 
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}. 
Results have been reported so far for basic planning algorithms like breadth-first 
search and \UCT\ \cite{bellemare:jair2013}, and for  reinforcement learning
and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the 
performance of  human players across a broad range of games.

@@

The presence of unknown transition and  rewards in  the Atari games does not preclude the use of blind-search
methods like breadth-first search, Dijkstra's algorithm \cite{dijkstra},  or learning methods  such as 
LRTA* \cite{korf:lrta}, \UCT\ \cite{uct}, and Q-learning  \cite{sutton:rl,bertsekas:neuro}. Indeed, 
the net-benefit planning problem with unknown state transitions and rewards over a given planning
horizon, can be mapped into a standard \emph{shortest-path problem} which can be solved optimally by Dijkstra's algorithm.
For this, we just need to  map the unknown rewards $r(a,s)$ into positive (unknown) action costs $c(a,s) = C - r(a,s)$
where  $C$ is a large constant that exceeds the maximum possible reward. The fact that the state 
transition and cost functions  $f(a,s)$ and $c(a,s)$ are not known a priori doesn't affect the applicability of 
Dijkstra's algorithm, which requires the value of these functions precisely when  the action $a$
is applied in the state $s$.

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line. 
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planning algorithms: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is 
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where 
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set) 
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one 
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and 
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is 
complete, while searching the state space in a way that makes use of the 
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables 
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$ 
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$, 
the procedure $\IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$. 
Elsewhere, it has been shown that  \IW\ exhibits   state-of-the-art performance on most existing planning benchmark domains 
when goals are restricted to be single atoms  where \IW\ provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a block world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time  
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that $\IW(2)$ will be 
not only complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for any instance, as long as the goal is atomic. 
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built, 
simple extensions of the basic \IW\ procedure have been developed as well, such as Serialized \IW, where \IW\ is 
used to achieve the goals one at at time \cite{nir:ecai2012,nir:ecai2014}. 
The extension of \IW\ for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

The Iterative Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the state space, a problem is defined by an initial state $s_0$, 
a set of actions applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not  need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached. 
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state 
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting, 
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW is the path  that has a  maximum accumulated reward. 
The accummulated reward $R(s)$  of a state $s$ reached in an iteration of \IW is determined by the  unique parent state $s'$ 
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is then the first action along such a path.

@@

\begin{table}[t!]
\begin{center}
%\resizebox{3in}{!}{
{\scriptsize
% \begin{tabular}{r@{}lrrrr@{}}
\begin{tabular}{llcccc}
%% table header
\hline\\[-1.3ex]
%\toprule
\# & \multicolumn{1}{c}{Domain} & \multicolumn{1}{c}{I} & \multicolumn{1}{c}{$\IW(1)$} & \multicolumn{1}{c}{$\IW(2)$} & \multicolumn{1}{c}{Neither} \\[.2ex]

@@

\caption{\small 
\emph{Top:} Number of classical planning problems  and percentages solved by $\IW(1)$, $\IW(2)$, or neither. Instances obtained
by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),  
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes and 2 GB resp. From \cite{nir:ecai2102}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very 
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition. 
Yet, this changes the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration 
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encodings are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$ 
\cite{bonet:hsp-aij} on the other (this algorithm is similar 
to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34627 vs. 34849, far ahead
of the other blind algorithms BrFS and ID  that solve 9010 and 8762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  
relevant  in  optimization problems as well. Indeed, if a  good path is made of 
states $s_i$ each of which have a low width, \IW\ can be made 
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property
which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case, 
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of 
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of 
such states. For classical planning problems, the growth in the number of nodes from 
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables, 
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and 
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't tried  this representation yet
under the assumption that the correlations among the bits in each one of 
the 128 words of the state vectors are meaningful. 
In summary, from the basic \IW\ algorithm we are testing only the first 
call $\IW(1)$.

\IW\ is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd,malte:alternating}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with no accumulated reward  that are 
placed in the first queue only. We refer to this best-first algorithm 
as \BFS.

For  the experiments below, we added two simple variations to  $\IW(1)$ and \BFS.
First, in the breadth-first search underlying $\IW(1)$, we generate the children in 
random order. This makes the executions that results from the
$\IW(1)$ lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms 
for discounting future rewards like in \UCT. 
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$, 
$R(s')$ is set to $R(s) + \gamma^d(s´) r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier, 
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

We tested  $\IW(1)$ and \BFS\  over $9$ games: those considered in the \emph{training} set
of \cite{bellemare:jair2013} and those discussed in \cite{deep-mind-atari}. 
In Table~\ref{performance} the performance of  $\IW(1)$ and \BFS\ 
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead, 
is compared with that of \UCT, breadth-first search (\BRFS), and random action selection. 
Our experimental setup  follows the one in  \cite{bellemare:jair2013} with two exceptions. First, rather than
considering $10$ episodes of each game, we have considered just $5$ due to time constraints, as the experiments 
for each  game can take up to $40$ hours of CPU time. Second, while \UCT\ is still limited to perform up to $500$ base policy rollouts 
of depth $300$, a maximum of budget of  $100,000$ simulated frames is applied to all the algorithms, \UCT\ included. 
%% Was not clear; hope this is what you meant, if not fix
%%% ($150,000$ frames are to be simulated while performing the lookahead), both $\IW(1)$ and \BFS are limited to
% The other \UCT\ parameters like the exploration constant $C$ and the horizon $m$ are the same ($C=0.1$, $m=300$). 
%% matching the setup used during actual gameplay. %%  (an action is selected every $5$ frames, using a no-op for the remaining $4$). %% Needed?
The average scores reported for \UCT\  \cite{bellemare:jair2013}, without this extra limitation, are included
in  parenthesis for reference. $\IW(1)$ and \BFS\ have been  limited to a planning horizon $m=1500$ where one an action is evaluated every 
$5$ frames. The discount factor used by all the algorithms is $\gamma=0.995$, and all of them  reuse the tree built 
in the previous lookahead by setting the root of the tree to the best child of the root, while 
deleting its siblings and their descendants as in \cite{bellemare:jair2013}.
%%% really need what's below? it's not very clear .. at least to me HG
%% We note that  this may have a different effect on algorithms where all branches have the same number of
%% generated nodes, as is the case of \BRFS, than in $\IW(1)$, \BFS\ and \UCT, where the depth of
%% the branches is uneven.
